---
title: "UFC Project"
author: Jared De La Serna
date: March 12, 2024
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

# Introduction:

UFC (Ultimate Fighting Championship) is a mixed martial arts (MMA)
organization where athletes must be well-rounded in various martial arts
disciplines. Competitors in the UFC come from various backgrounds such
as Brazilian jiu-jitsu, boxing, wrestling, judo, karate, Muay Thai, and
more. Being the best in the UFC often requires proficiency in both
striking and grappling techniques, as well as excellent conditioning,
perfect timing, and mental strength. This diverse skill set allows
fighters to handle the wide range of combat scenarios they encounter in
the octagon. UFC Championship is divided into 8 weight divisions, weight
is the most crucial in the sport, you have to make a certain weight a
day before the fight is one of the most important day before the fight,
failure to meet the weight requirements you can be fine, titled will
vacant, or the fight can be canceled.

We will explore this UFC dataset that has statistics of all the UFC
fighters. I got the dataset from Kaggle, which data of 2,479 fighters,
and it had statistics from every aspect of the sport.

The variables are: 1) SLpM - Significant Strikes Landed per Minute

2)  Str. Acc. - Significant Striking Accuracy

3)  SApM - Significant Strikes Absorbed per Minute

4)  Str. Def. - Significant Strike Defence (the % of opponents strikes
    that did not land)

5)  TD Avg. - Average Takedowns Landed per 15 minutes

6)  TD Acc. - Takedown Accuracy

7)  TD Def. - Takedown Defense (the % of opponents TD attempts that did
    not land)

8)  Sub. Avg. - Average Submissions Attempted per 15 minutes

# Method Description

Imputation For Missing Values

# Research Question

How can we input values from fights that happened but the statistics of
the fight were never recorded for reasons such as; the fight not being
televised, UFC not posting the actual statistics for the fight, or data
collection issues?

# Descirption of the Method:

A technique used to handle missing data. The key idea is to create
several different plausible imputations for the missing values, rather
than filling them with a single estimate like the mean or median. This
has many purposes, it could be that you can either ignore the NA's
valuesi in your dataset, or you can imput them based on predictive
factors.

# Motivation of Method:

Within the UFC there's been a long history of unrecorded fights within
the entire history of the organization. This makes it difficult to track
and understand how the sport has been involved over time. I really want
to see what would happen if the UFC would release the true statistics of
the fighter, because most of them are not televised or not enough
importance in order to publish their statistics. Which I think is unfair
to have the them as N/A.

# Small Example

Since the imputation for missing values is complicated, I'll demonstrate
the power of the method with a smaller dataset. The example was from
"Imputation For Missing Values (HSUAR Chapter 16)" by Torsten Hothorn
and Brian S. Everitt.

We create the dataset manually by imputing and creating all the
variables. We will show the dataset. This dataset records lowest
temperatures (in Fahrenheit) recorded in various months for cities in
the US. Warning: there's some NA's.

```{r}
temperature_data <- data.frame(
  City = c("Atlanta", "Baltimore", "Bismarck", "Boston", "Chicago", "Dallas", "Denver", "El Paso",
           "Honolulu", "Houston", "Juneau", "Los Angeles", "Miami", "Nashville", "New York",
           "Omaha", "Phoenix", "Portland", "Reno", "San Francisco", "Seattle", "Washington"),
  January = c(-8, -7, -44, -12, -27, 4, -25, -8, 53, 12, -22, 23, 30, -17, -6, -23, NA, -26, -16, 24, NA, -5),
  April = c(26, 20, -12, 16, 7, NA, -2, 23, 57, 31, 6, 39, 46, 23, 12, 5, 32, 8, NA, 31, 29, 24),
  July = c(53, NA, 35, 54, 40, 59, 43, 57, 67, 62, 36, 49, 69, 51, 52, 44, 61, 40, 33, 43, 43, 55),
  October = c(28, 25, 5, 28, 17, 29, 3, NA, NA, 33, 11, NA, 51, 31, 14, 13, 34, 15, 8, NA, 28, 29)
)

print(temperature_data)

```

We can see our dataset has some NA that might hurt the the credibility
of dataset. In reality we have two options, either get rid of the NA's
or we can use Imputation For Missing Values in order to replace our NA's
with a value. You can either do both, but for the power of learning we
will impute values for the missing values.

First step, we try to identify on our NA's in our dataset. This could be
done with using the "sapply" function and doing the a sum() in order to
count all the NA's.

```{r}
sapply(temperature_data, function(x) sum(is.na(x)))
```

We see have multiple missing values in the months. Now we will try to
imput values for October since it's the month that has the most missing
values. Note you can do this for any variables.

First we create a dataset called "imp" which will contain all variables;
City, January, April, July ,and October. The missing values in October
will be replaced by the mean temperatures in October of the complete
cases. We will talk about the package and library "mice" later on.

```{r}
library(mice)
imp <- mice(temperature_data, method = "mean", m = 1, maxit = 1)
```

Now we can to compare our summary statistics of the IMP and our unedited
dataset.

```{r}
summary(temperature_data$October)
with(imp, summary(October))
```

We can see the that the values of the 1st Quartile, median, and 3rd
Quartile has changed, but the mean has remained unchanged. Another
useful fact is that we get to see how many missing values are in the
dataset. Also usually, the mean won't change because the missing values
that we had we replaced by the true sample mean. Which some cases when
it comes to data cleaning that might be the right thing to, but in this
problem we might have to do it differently.

Now we want to see if there's a change in standard deviation.

```{r}
sd(temperature_data$October, na.rm = TRUE)
with(imp, sd(October))
```

We can see that the standard deviation is 12.3336 just ignoring the NA's
of our raw dataset. It's 11.09698 when we input the average mean and
replace them with the NA's dataset. This means the values for October
are clustered tightly around the mean more with the imputation of mean
instead of having NA's. Which it might be a good method to replace our
values, but we need to dive deeper into the analysis. What if we find a
better way to do it?

First of all lets see if we can see a correlation between all our
variables when comparing it to October.

```{r}
layout(matrix(1:3, nrow = 1))
plot(October ~ January, data = temperature_data)
plot(October ~ April, data = temperature_data)
plot(October ~ July, data = temperature_data)
legend("topleft", pch = 1:2, bty = "n",
legend = c("original", "imputed"))
```

We might see as the temperature increases in the months of January,
April, and July the temperature in October seems to be going up. We see
a positive correlation between the temperature of October compare to the
temperature of other months.

Now we will show how those imputed values look like in a scatter plot,
we expect to see all the mean values that are imputed to be in a
horizontal line. We will compare the correlation between October to
other months.

```{r}
layout(matrix(1:3, nrow = 1))
plot(October ~ January, data = complete(imp),
pch = is.na(temperature_data$October) + 1)
plot(October ~ April, data = complete(imp),
pch = is.na(temperature_data$October) + 1)
plot(October ~ July, data = complete(imp),
pch = is.na(temperature_data$October) + 1)
legend("topleft", pch = 1:2, bty = "n",
legend = c("original", "imputed"))
```

As you can see all the imputed values are the mean of the temperate of
October. As we can see this method of imputing value might not be the
best approach to accurately input values based on many factors. We can
see that we are suggesting that all the values that were imputed are the
same in October no-matter what month we compare to such as January,
April, or July. Now assuming that the months have correlation between
each other, which sometimes they do when we look at the correlation
between temperature and month. The imputed values don't follow the
pattern that we see, so we might have to use a different method to
impute values.

Now we will use "predictive mean matching", which is known as method =
"pmm" which basically predicts values close to the missing value's
prediction for imputation. This could be usually when we see a
correlation between our variables, and the imputed value from the pmm
will imput a value based on the pattern that they detec in the
algorithm. We are doing 20 iterations, because our sample size is close
to 20.

```{r}
imp_ppm <- mice(temperature_data, m = 20, method = "pmm",
print = FALSE, seed = 1)
```

We will graph the same plots but now we will the the "ppm" method in
order to see how it works differently compared to "imp" method.

```{r}
layout(matrix(1:3, nrow = 1))
plot(October ~ January, data = complete(imp_ppm),
pch = is.na(temperature_data$October) + 1)
plot(October ~ April, data = complete(imp_ppm),
pch = is.na(temperature_data$October) + 1)
plot(October ~ July, data = complete(imp_ppm),
pch = is.na(temperature_data$October) + 1)
legend("topleft", pch = 1:2, bty = "n",
legend = c("original", "imputed"))
```

We can see there are imputed values that follow the trends of the
correlations of temperature between months, which can make us assume
that we can use values for those values instead of NA's, but we just
have to disclose that we put those missing values based on our "imp_ppm"
method. We can say that when we use imp_ppm method we tend those values
that were imputed follow the correlation that say before doing any
method. This might be a better than method than predictive mean
matching.

```{r}
summary(unlist(with(imp_ppm, sd(October))$analyses))
```

Right here it summarizes the standard deviations of the October variable
across multiple imputed data sets in imp_ppm. As we can see have a
minimum of 11.30 and a maximum of 12.85 even though our standard
deviation might not be the lowest compared to imp method. It makes sense
because imp was just imputing straight mean values which are totally
inaccurate, these standard deviations have to vary because every value
imputed is based on predictive mean matching.

Now we just will do some quick t.test in order to compare how each test
vary.

```{r}
with(temperature_data, t.test(October, mu = 20))
with(imp, t.test(October, mu = 20))$analyses[[1]]
fit <- with(imp_ppm, lm(I(October - 20) ~ 1))
summary(pool(fit))
```

We can see that when we test, in our case, the p-values are above 0.05,
meaning there isn't strong evidence to reject the null hypothesis that
the true mean temperature in October is 20°C. We can see how every model
is different that they provide different p-test. Depending what your
test and research consint we should be able to pick what method we think
is more effective.

# Read and Load Data

To conduct Imputation For Missing Values in R, it is important we
install and load the package and the library mice. The mice package
implements a method to deal with missing data. The package creates
multiple imputations for multivariate missing data.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mice)
library(dplyr)
library(readr)
library(ggplot2)
library(corrplot)
```

```{r}
fighter_stats <- read_csv("fighter_stats (1).csv")
```

# Exploratory Data Analysis

We don't have a column for weight division, we decide to mutate in order
to make columns, because every division has a certain fighting division,
we made created a new column, also the weight was in kg so we coveted
into lbs and made sure every fighter was into their right category.

```{r echo=FALSE}
fighter_stats <- fighter_stats |>
  mutate(`Weight(lbs)` = weight * 2.20462) |>  
  filter(`Weight(lbs)` >= 115) |>  
  mutate(`Weight_Class` = case_when(
    `Weight(lbs)` < 125                        ~ "Flyweight",
    `Weight(lbs)` >= 125 & `Weight(lbs)` < 135 ~ "Bantamweight",
    `Weight(lbs)` >= 135 & `Weight(lbs)` < 145 ~ "Featherweight",
    `Weight(lbs)` >= 145 & `Weight(lbs)` < 155 ~ "Lightweight",
    `Weight(lbs)` >= 155 & `Weight(lbs)` < 170 ~ "Welterweight",
    `Weight(lbs)` >= 170 & `Weight(lbs)` < 185 ~ "Middleweight",
    `Weight(lbs)` >= 185 & `Weight(lbs)` < 205 ~ "Light Heavyweight",
    `Weight(lbs)` >= 205 & `Weight(lbs)` < 265 ~ "Heavyweight",
    `Weight(lbs)` >= 265                      ~ "Super Heavyweight"
  ))
```

We want to rename our variables in order to better understanding of our
variables. Names might get confusing.

```{r echo=FALSE}
fighter_stats <- fighter_stats %>%
  rename(
    `Significant_Strikes_Landed_per_Minute` = SLpM,
    `Significant_Striking_Accuracy_` = sig_str_acc,
    `Significant_Strikes_Absorbed_per_Minute` = SApM,
    `Significant_Strike_Defence_` = str_def,
    `Average_Takedowns_Landed_per_15_minutes` = td_avg,
    `Takedown_Accuracy` = td_acc,
    `Takedown_Defense` = td_def,
    `Average_Submissions_Attempted_per_15_minutes` = sub_avg
  )
```

# We want to see a comparison between Reach and Weight Class.

```{r echo=FALSE}
ggplot(fighter_stats, aes(x = reach, y = Significant_Strikes_Landed_per_Minute, color = Weight_Class)) +
  geom_point(alpha = 0.5) + 
  labs(title = "Scatter Plot: Reach vs. Weight Class",
       x = "Reach (inches)", y = "Significant Strikes Per Minute (SLpM)") +
  theme_minimal() 

```

We can see a normal distributed data.We can see that even though we
would think that there's a correlation between weight_class and
significant strikes per minutes. We can say this because striking is one
of the biggest attributes when it comes to fighting in the UFC. But we
can see a correlation between weight classes and reach, and which make
sense the more weight a fighter carries, they tend to be taller and
being taller leads to a much longer reach.

# A correlation matrix between all numeric variables.

```{r echo=FALSE}
cor_matrix <- cor(fighter_stats[, sapply(fighter_stats, is.numeric)], use = "complete.obs")
corrplot(cor_matrix, method="color", tl.cex=0.6) 
```

the correlation plot shows a strong positive correlation between a
fighter's weight and height. This indicates that taller fighters tend to
be heavier, which is consistent. There is a positive correlation between
offensive performance metrics such as significant strikes landed per
minute and average takedowns landed per 15 minutes. This suggests that
fighters who are more aggressive in striking are also more likely to
attempt and succeed in takedowns, which can show an overall aggressive
fighting style. The plot reveals correlations between defensive metrics
such as strike defense and takedown defense. This implies that fighters
who excel at defending against strikes also tend to be better at
avoiding or countering takedowns. We can see more correlations.

# Distribution of Weight Classes

```{r echo=FALSE}
ggplot(fighter_stats, aes(x = Weight_Class)) +  #
  geom_bar(stat = "count", fill = "maroon", color = "black") + 
  labs(title = "Distribution of Weight Classes",
       x = "Weight Classes",  
       y = "Frequency") + 
  theme_minimal() 

```

We can see the frequency across every weight division, the amount of
fighters that there is the most popular weight division is welterweight
and super heavyweight is the least popular weight decision in terms of
number of fighters. The UFC divisions are mostly have the same watch
popularity which is good for the buisness, but usually when we have
light heavyweight and super heavy wight fights, since the amount of
fighters in that division is so low, usually those fights are the most
watched.

# Reach vs Height

```{r echo=FALSE}
ggplot(fighter_stats, aes(x = height, y = reach)) +
  geom_point(aes(color = Weight_Class), alpha = 0.5) + 
  labs(title = "Reach vs. Height",
       x = "Height (cm)", 
       y = "Reach (cm)") + 
  theme_minimal() 

```

We can see that the strong positive relationship between height and
reach. Which supports what our matrix plot shows.

# Average Submission Average by Age and Weight Class

```{r}
ggplot(fighter_stats, aes(x = age, y = Average_Submissions_Attempted_per_15_minutes, fill = Weight_Class)) +
  geom_col() + 
  labs(title = "Average Submission Average by Age and Weight Class",
       x = "Age",
       y = "Average Submissions Attempted per 15 Minutes") +  # Corrected label description
  theme_minimal() 
```

We see that ages are evenly distributed through ages and how we can see
the how in sporting events there's a term where is prime which can be
consider when its around 25-35 who they are people who have the highest
average reach making them strongest and most aggressive.We can see a
pattern between all these 3 variables, we can see how age and the
average submission are correlated and how all the weight classes tend to
follow the same correlation.

# Data Science Method: Imputation For Missing Values

We noticed that our dataset had many missing variables. This is mainly
because we UFC started they were really counting significant strikes or
televised certain fights so their statistics become unknown. We want
those numbers as NA in order to impute the missing values based on other
variables.

Replace all the zeroes in your dataset with NA (Not Available)

```{r}
fighter_stats[fighter_stats == 0] <- NA
```

Counting how many NA's we have in our dataset using a sapply() and
seeing how many NA have per variable.

```{r}
na_count <- sapply(fighter_stats, function(x) sum(is.na(x)))
print(na_count)
```

The variables I want to impute are reach because we don't have any
missing values for height so predicting each UFC fighter can be
relatively easy. I also want to input values for reach because reach is
something that I feel can have strong correlation with striking and
height. The data about reach of a fighter shouldn't be NA because every
fighter has a reach no matter how tall or short they are. 645 missing
values for reach is concerning.

# Exploratory data analysis for the 'reach' variable:

```{r}
library(mice)

layout(matrix(1:3, nrow = 1))
plot(reach ~ height, data = fighter_stats)
plot(reach ~ `Weight(lbs)`, data = fighter_stats)
plot(reach ~ Significant_Strikes_Landed_per_Minute, data = fighter_stats)

```

The scatter plots display the relationships between fighters' reach and
their height, weight, and significant strikes landed per minute. We can
see that there is a positive correlation between height and reach, as
well as between weight and reach, suggesting that taller and heavier
fighters generally have a longer reach. But there is not to be a clear
relationship between reach and the rate of significant strikes landed
per minute.

# Linear model summary for 'reach':

```{r}
summary(lm(reach ~ height + `Weight(lbs)` + Significant_Strikes_Landed_per_Minute, data = fighter_stats))
```

We can see that p value for intercept, height, and weight, is
significant, but its not significant with Significant Strikes Landed per
Minute which totally correlates with our scatter plots.

Now as we did in the small example before we are going to do "imp"
again. Which is single imputation using mean values:

```{r}
imp <- mice(fighter_stats, method = "mean", m = 1, maxit = 1)
with(imp, summary(reach))
with(imp, sd(reach))
with(imp, cor(height, reach))
```

First of all we can see a strong correlation between height and reach
which is around .74485, which makes sense. We can see a standard
deviation of 8.402345 which is significantly high. More research is
needed. Now let see our mean imputed values and see how does it look in
a plot.

# Visualizing imputed data:

```{r}
layout(matrix(1:3, nrow = 1))
plot(reach ~ height, data = complete(imp),
pch = is.na(fighter_stats$reach) + 1)
plot(reach ~ `Weight(lbs)`, data = complete(imp),
pch = is.na(fighter_stats$reach) + 1)
plot(reach ~ `Significant_Strikes_Landed_per_Minute`, data = complete(imp),
pch = is.na(fighter_stats$reach) + 1)
legend("topright", pch = 1:2, bty = "n",
legend = c("original", "imputed"))
```

We see that the values that were imputed are the means of reach, which
it's nice visualization as we can we see where the missing values are
really located. But in reality, it just a horizontal line of imputed
value which makes sense because its the average of reach imputed across
height,weight, and significant strikes landed per minute. But we need to
do predictive mean matching in order to determine which one is better.

Now we use Multiple Imputation using predictive mean matching, we do 50
iterations, compared to 20 in the example.

```{r}
imp_ppm <- mice(fighter_stats, m = 50, method = "pmm",
print = FALSE, seed = 1)
```

Visualizing results from multiple imputation:

```{r}
layout(matrix(1:3,nrow = 1))
plot(reach ~ height, data= complete(imp_ppm),
pch = is.na(fighter_stats$reach) + 1)
plot(reach ~ `Weight(lbs)`, data= complete(imp_ppm),
pch = is.na(fighter_stats$reach) + 1)
plot(reach ~ `Significant_Strikes_Landed_per_Minute`, data = complete(imp_ppm),
pch = is.na(fighter_stats$reach) + 1)
legend("topright", pch = 1:2, bty = "n",
       legend = c("orginal", "imputed"))
```

As we can see the values that were imputed mostly follow the pattern
that it does when its two variables, which can help us predict in a way
what would happen if the actual statistics were imputed. Since UFC is a
really unpredictable sport, this could give us a sense of what would a
fighter would have been if we didn't know the actual statistics. Since
weights and height have a strong correlation with reach we have to use
this model better.

# Summary statistics of multiple imputation results:

```{r}
summary(unlist(with(imp_ppm, sd(reach))$analyses))
summary(unlist(with(imp_ppm,
    cor(height, reach))$analyses))
summary(unlist(with(imp_ppm,
    cor(`Weight(lbs)`, reach))$analyses))
sd(fighter_stats$reach, na.rm = TRUE)
```

We can see that our standard deviation might be higher than the actual
standard deviation, but that is fine because as mentioned earlier, UFC
is one of the most unpredictable sports, because there are so many
factors that play into fight day such as adrenaline, timing, and
training. The purpose of mice is not to improve the standard deviation
of our dataset but instead to impute missing values making sure our
dataset doesn't include NA. Indeed, adding predicted values will usually
increase the standard deviation but that's not always the case. Now we
will do a T test in order to see how testing with a certain mean value.

Now we test our models without imp, with imp_pp, and with imp_ppm. We
should see a difference in the p-value.

```{r}
t.test(fighter_stats$reach, mu =183)
with(imp, t.test(reach, mu = 183))$analyses[[1]]
with(imp_ppm, t.test(reach, mu = 183))$analyses[[1]]
```

Imp_ppm likely provided a more accurate and varied estimation of the
missing data. Multiple imputation generally produces more reliable
estimates because it accounts for the uncertainty in the imputation
process by creating several different imputations and pooling the
results. We can also conclude that models can perform more than others,
there might be instances where a model with just the mean of the
variable imputed might be different, and there might be instances where
the predictive mean matching might be better. It all depends on the
context of the dataset. As we can our p-value was less than 0.05 for
imp_ppm, while in the other hand normally and imp were relative the
same, and way bigger than 0.05. Which this why for imputing values for
reach, I would use imp_ppm.

# Conclusion

I believe if you have missing values, and you have to impute them and
the variable that has missing values is not correlated to your dataset I
will use imp, but if you have strong correlation between all your
variables across I would use imp_ppm. It all to the researcher
discretion.

# Bibliography

@misc{everitt2013hsaur, title={HSAUR: a handbook of statistical analyses
using R. R package}, author={Everitt, BS and Hothorn, T}, year={2013} }

\`\`\`
